### 2025-01-24 paper/icml25/first/sects/background.tex
few corrections...

\subsection{Markov Decision Processes}
To provide context for our work, we first introduce the reinforcement learning (RL) framework. Each task is modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$. Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively. $\allo=\{\observations{i}\}_{i\in\agents}$ and $\alla=\{\actions{i}\}_{i\in\agents}$ are the observation space and action space for each agent. $\{\stateobs{i}:\states\rightarrow\observations{i}\}_{i\in\agents}$ are the projections from the state to the observation space and $\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}_{i\in\agents}$ are the reward functions for each agent. $\transition:\states\times\alla$ denotes the transition function. $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively. Finally, $\initdist$ is the initial state distribution.

Further, let us denote denote the space of trajectories, $\trajectories = \left(\states\times\alla\times\states\times\mathbb{R}^{n}\right)^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent. Given a generic trajectory $\tau\in\trajectories$, we denote its $i$-th starting state with $s_i^\tau$, its $i$-th action for agent $j$ with $a_{ij}^\tau$, and its $i$-th reward for agent $j$ with $r_{ij}^\tau$.

@@

The goal of the MDP is to find a policy that maximizes the expected return 
$$\mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{\horizon-1}\sum_{i\in\agents}\gamma^t r_{ti}^\tau \right]$$

@@

In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled from $\mdp$. The loss $\mathcal{L}_\theta^{\pi}(\batch)$ is defined as:  
\begin{equation*}\label{eq:shac}
    \frac{1}{\bs\na\mh}\sum_{\tau\in\batch} \sum_{t=t_0}^{t_0+\mh-1} \sum_{i\in\agents} \discount^{t-t_0} R(s_{t}^\tau, a_{ti}^\tau) + \discount^\mh \valuef_\theta(s^\tau_{t_0+\mh})
\end{equation*}
where $\mh \ll \horizon$ is the roullout size, $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return. It is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$. While this is a common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

>>>>

\subsection{Markov Decision Processes}
To provide context for our work, we first introduce the reinforcement learning (RL) framework. Each task is modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$. Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively. $\allo={\{\observations{i}\}}_{i\in\agents}$ and $\alla={\{\actions{i}\}}_{i\in\agents}$ are the observation space and action space for each agent. ${\{\stateobs{i}:\states\rightarrow\observations{i}\}}_{i\in\agents}$ are the projections from the state to the observation space and ${\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}}_{i\in\agents}$ are the reward functions for each agent. $\transition:\states\times\alla$ denotes the transition function. $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively. Finally, $\initdist$ is the initial state distribution.

Further, let us denote denote the space of trajectories, $\trajectories = {\left(\states\times\alla\times\states\times\mathbb{R}^{n}\right)}^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent. Given a generic trajectory $\tau\in\trajectories$, we denote its $i$-th starting state with $s_i^\tau$, its $i$-th action for agent $j$ with $a_{ij}^\tau$, and its $i$-th reward for agent $j$ with $r_{ij}^\tau$.

@@

The goal of the MDP is to find a policy that maximizes the expected return 
\[\mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{\horizon-1}\sum_{i\in\agents}\gamma^t r_{ti}^\tau \right]\]

@@

In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled from $\mdp$. The loss $\mathcal{L}_\theta^{\pi}(\batch)$ is defined as:  
\begin{equation*}\label{eq:shac}
    \frac{1}{\bs\na\mh}\sum_{\tau\in\batch} \sum_{t=t_0}^{t_0+\mh-1} \sum_{i\in\agents} \discount^{t-t_0} R(s_{t}^\tau, a_{ti}^\tau) + \discount^\mh \valuef_\theta(s^\tau_{t_0+\mh})
\end{equation*}
where $\mh \ll \horizon$ is the rollout size, $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return. It is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$. While this is a common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

### 2025-01-24 paper/icml25/first/sects/discussion.tex
few corrections...

\begin{center}
\RQ{2} \textbf{How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?}
\end{center}
The proposed algorithm appears capable of outperforming PPO across all scenarios showing also a better sample efficiency. In single-agent cases, while PPO manages to find near-good policies (e.g., in the Dispersion scenario), it often fails to complete scenarios within the same number of steps. Indeed, \fname{} achieves better results in all environments (see \Cref{tab:max-rewards}). Despite having to train multiple networks, \fname{} converges much faster to optimal policies, as gradient guidance enables quicker and more accurate convergence to the desired outcome. This pattern becomes even more evident in multi-agent scenarios. In these cases, PPO often struggles to find even marginally good policies. \fname{}, on the other hand, eventually succeeds in understanding the environment and subsequently improves the policy to enable cooperation---see both \Cref{tab:max-rewards} and \Cref{fig:experiments}.

\begin{center}
\RQ{3} \textbf{How does the performance of these algorithms change as the search space increases?}
\end{center}
To analyze the impact of increasing search spaces, we primarily focused on scenarios with varying numbers of agents. The results demonstrate that performance patterns differ across scenarios. In the Dispersion scenario, \fname{} shows increasing difficulty in finding optimal policies as the number of agents grows. Conversely, in Transport and Sampling scenarios, the system appears to converge more quickly with more agents. This counterintuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward. Even with approximated world dynamics (transition and reward functions), the guidance provided is sufficient to lead policies toward optimal behavior.

This advantage does not hold in the Dispersion scenario, where complexity increases substantially with additional agents. Nevertheless, we observe the emergence of cooperation across all scenarios, suggesting potential scalability to larger agent populations. The main bottleneck lies in the transformer architecture for the action world model, where the sequence length scales with the number of agents (sequence length = num. agents times num. train steps). With significantly large populations (~50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively large memory requirements. In these scenarios, a linear attention mechanism (such as \cite{Beltagy20}) may be employed.

>>>>

\begin{center}
\RQ{2} \textbf{How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?}
\end{center}
The proposed algorithm appears capable of outperforming PPO across all scenarios showing also a better sample efficiency. In single-agent cases, while PPO manages to find near-good policies (e.g., in the Dispersion scenario), it often fails to complete scenarios within the same number of steps. Indeed, \fname{} achieves better results in all environments (see \Cref{tab:max-rewards}). Despite having to train multiple networks, \fname{} converges much faster to optimal policies, as gradient guidance enables quicker and more accurate convergence to the desired outcome. This pattern becomes even more evident in multi-agent scenarios. In these cases, PPO often struggles to find even marginally good policies.\ \fname{}, on the other hand, eventually succeeds in understanding the environment and subsequently improves the policy to enable cooperation---see both \Cref{tab:max-rewards} and \Cref{fig:experiments}.

\begin{center}
\RQ{3} \textbf{How does the performance of these algorithms change as the search space increases?}
\end{center}
To analyze the impact of increasing search spaces, we primarily focused on scenarios with varying numbers of agents. The results demonstrate that performance patterns differ across scenarios. In the Dispersion scenario, \fname{} shows increasing difficulty in finding optimal policies as the number of agents grows. Conversely, in Transport and Sampling scenarios, the system appears to converge more quickly with more agents. This counter intuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward. Even with approximated world dynamics (transition and reward functions), the guidance provided is sufficient to lead policies toward optimal behavior.

This advantage does not hold in the Dispersion scenario, where complexity increases substantially with additional agents. Nevertheless, we observe the emergence of cooperation across all scenarios, suggesting potential scalability to larger agent populations. The main bottleneck lies in the transformer architecture for the action world model, where the sequence length scales with the number of agents (sequence length = num.\ agents times num.\ train steps). With significantly large populations (~50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively large memory requirements. In these scenarios, a linear attention mechanism (such as~\cite{Beltagy20}) may be employed.

### 2025-01-24 paper/icml25/first/sects/experiments.tex
minors
few corrections...

\begin{figure*}[!t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}
    \label{fig:experiments}
\end{figure*}

\begin{figure}[t]
    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Norm of policy gradients on the Transport environment across 5000 epochs. The raise in norm for SHAC (epoch 2700) corresponds with occurence of generalization, where collision between agents and the packages become frequent. Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
\end{figure}

@@

\paragraph{Scenarios}
As mentioned previously, we choose a few multi-agent scenarios from the VMAS package to test \fname{}, SHAC and PPO. In particular, we choose the scenarios Dispersion, Discovery, Transport, and Sampling. For details regarding each scenario, see \Cref{apx:scenarios}. Dispersion and Discovery have non-differentiable reward functions, thus SHAC is not applicable. Transport and Sampling have differentiable reward functions, thus SHAC is applicable. For Transport and Dispersion, the observation space is complete wrt. the state. For Discovery and Sampling, the observation space is incomplete wrt. the state. In these scenario, \fname{} is slightly disadvantaged as it does not have access to the full state of the environment compared to SHAC.

@@

\paragraph{Hardware Setup}
Our experiments were conducted on a cluster of machines composed of 2 computational nodes, each with one Nvidia Tesla V100 GPU (with 32GB of memory), 200GB of RAM, and two Intel Xeon Gold 6226R processors (32 cores each).

@@

\paragraph{Transport}
The third row of \Cref{fig:experiments} presents a differentiable scenario, transport. In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance. \fname{} converges faster, but SHAC is more consistent due to direct gradient stability. In the second and third columns (3 and 5 agents), PPO fails as in the single-agent case, while SHAC and \fname{} both complete the task around 10,000 episodes. With more agents, convergence is faster because the task becomes easier (as more agents can push the package). The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package. This is a sign of generalization, as the agents learn to avoid collisions and push the package together.

\paragraph{Sampling}
The fourth row of \Cref{fig:experiments} illustrates the sampling scenario, which is differentiable and thus applicable to SHAC. In the first column (single agent), PPO fails to converge, while \fname{} outperforms SHAC. In the second and third columns (3 and 5 agents), PPO again fails to converge, while SHAC and \fname{} both complete training in fewer than 5,000 episodes. Notably, SHAC does not exhibit the same stability advantage it displayed in other scenarios, possibly because gradients derived directly from the environment are noisier compared to those learned by the transition network.

>>>>

\begin{figure*}[!t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
    \label{fig:experiments}
\end{figure*}

\begin{figure}[t]
    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Norm of policy gradients on the Transport environment across 5000 epochs. The raise in norm for SHAC (epoch 2700) corresponds with occurrence of generalization, where collision between agents and the packages become frequent. Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
\end{figure}

@@

\paragraph{Scenarios}
As mentioned previously, we choose a few multi-agent scenarios from the VMAS package to test \fname{}, SHAC and PPO\@. In particular, we choose the scenarios Dispersion, Discovery, Transport, and Sampling. For details regarding each scenario, see \Cref{apx:scenarios}. Dispersion and Discovery have non-differentiable reward functions, thus SHAC is not applicable. Transport and Sampling have differentiable reward functions, thus SHAC is applicable. For Transport and Dispersion, the observation space is complete wrt.\ the state. For Discovery and Sampling, the observation space is incomplete wrt.\ the state. In these scenario, \fname{} is slightly disadvantaged as it does not have access to the full state of the environment compared to SHAC\@.

@@

\paragraph{Hardware Setup}
Our experiments were conducted on a cluster of machines composed of 2 computational nodes, each with one NVIDIA Tesla V100 GPU (with 32GB of memory), 200GB of RAM, and two Intel Xeon Gold 6226R processors (32 cores each).

@@

\paragraph{Transport}
The third row of \Cref{fig:experiments} presents a differentiable scenario, transport. In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance.\ \fname{} converges faster, but SHAC is more consistent. In the second and third columns (3 and 5 agents), PPO fails as in the single-agent case, while SHAC and \fname{} both complete the task around 10,000 episodes. With more agents, convergence is faster because the task becomes easier (as more agents can push the package). The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package. This is a sign of generalization, as the agents learn to avoid collisions and push the package together.

\paragraph{Sampling}
The fourth row of \Cref{fig:experiments} illustrates the sampling scenario, which is differentiable and thus applicable to SHAC\@. In the first column (single agent), PPO fails to converge, while \fname{} outperforms SHAC\@. In the second and third columns (3 and 5 agents), PPO again fails to converge, while SHAC and \fname{} both complete training in fewer than 5,000 episodes. Notably, SHAC does not exhibit the same stability advantage it displayed in other scenarios, possibly because gradients derived directly from the environment are noisier compared to those learned by the transition network.

### 2025-01-24 paper/icml25/first/sects/introduction.tex
few corrections...

\section{Introduction}\label{sect:introduction}

Learning control policies has important applications in robotics~\cite{Singh22}, autonomous driving~\cite{Elallid22}, swarm intelligence~\cite{Tang21} and many other fields. These scenarios present a variety of different challenges, such as sparse rewards, complex dynamics, and multi-agent interactions. Current approaches often focus on either single- or multi-agent scenarios, lacking a unified solution that works effectively in both settings. Moreover, reinforcement learning algorithms tend to be \emph{sample inefficient}, namely requiring a large number of interactions with the environment to learn a good policy.
%\texttt{\color{DarkOrchid}{<- FB: non mi è super chiara questa frase}}. 
A recently introduced method for the latter problem is SHAC~\cite{Xu22} which exploits differentiable simulators to provide precise gradient information, thereby expediting policy optimization and improving learning efficiency.

@@

We summarize our contributions as follows:
\begin{compactitem}
    \item We propose a new reinforcement learning framework that lift the requirement for a differentiable simulator in SHAC.
    \item We assess the SHAC algorithm in multi-agent scenarios, where it has not been previously evaluated.
    \item We evaluate our algorithm against PPO and SHAC on a set of environments with increasing numbers of agents.
    \item We show that our algorithm achieve SHAC-like performance in differentiable environments and outperform PPO in non-differentiable environments.
\end{compactitem}

>>>>

\section{Introduction}\label{sect:introduction}
Learning control policies has important applications in robotics~\cite{Singh22}, autonomous driving~\cite{Elallid22}, swarm intelligence~\cite{Tang21} and many other fields. These scenarios present a variety of different challenges, such as sparse rewards, complex dynamics, and multi-agent interactions. Current approaches often focus on either single- or multi-agent scenarios, lacking a unified solution that works effectively in both settings. Moreover, reinforcement learning algorithms tend to be \emph{sample inefficient}, namely requiring a large number of interactions with the environment to learn a good policy.
A recently introduced method for the latter problem is SHAC~\cite{Xu22} which exploits differentiable simulators to provide precise gradient information, thereby expediting policy optimization and improving learning efficiency.

@@

We summarize our contributions as follows:
\begin{compactitem}
    \item We propose a new reinforcement learning framework that lift the requirement for a differentiable simulator in SHAC\@.
    \item We assess the SHAC algorithm in multi-agent scenarios, where it has not been previously evaluated.
    \item We evaluate our algorithm against PPO and SHAC on a set of environments with increasing numbers of agents.
    \item We show that our algorithm achieve SHAC-like performance in differentiable environments and outperform PPO in non-differentiable environments.
\end{compactitem}
